---
title: "assignment 3"
format: html
editor: visual
---

## Methodology

This project investigates the impact of social media addiction on students using a publicly available dataset. Key variables include daily usage hours, addiction scores, academic performance impact, mental health ratings, interpersonal conflicts, and most used platforms. Data cleaning involved checking for missing values.

Descriptive and exploratory analyses were conducted to examine associations between addiction levels and outcomes. Boxplots, bar charts, and scatter plots were created to visualize variable relationships. A summary table was also generated to compare average usage hours and addiction scores by platform.

### Dataset Description

The student social media addiction data has `r nrow(addiction)` observations and `r ncol(addiction)` variables.

```{r}
library(tidyverse)
```

```{r}
addiction <- read.csv("data/Students Social Media Addiction.csv")
```

```{r}
head(addiction,2)
```

```{r}
#| label: tbl-variabletable
#| tbl-cap: Variable Names of the addiction Data

missing_data <- data.frame(
  Missing_Count = colSums(is.na(addiction))
)
missing_data
```

@tbl-variabletable shows the missing values across all variables. A summary table was generated to count missing entries per column. As shown in Table X, all key variables such as addiction score, academic performance impact, sleep hours, and mental health score had no missing values, indicating a complete dataset suitable for analysis without imputation.

```{r}
#| label: fig-Distribution
#| fig-cap: "Distribution of Addicted Scores"
#| eval: true

hist(addiction$Addicted_Score,
     breaks = 6,
     col = "skyblue",
     border = "black",
     main = "Distribution of Addicted Scores",
     xlab = "Addicted Score",
     ylab = "Frequency")
```

@fig-Distribution illustrates the distribution of self-reported addiction scores (1–10) among all participants. The majority of students fall in the 6–8 range, with the highest bar at 6–7, indicating a sample skewed toward moderate-to-high addiction levels. Scores below 4 are comparatively rare. This left-skewed distribution confirms that Addicted_Score is a meaningful and varied independent variable for our subsequent descriptive analyses.