---
title: "assignment 3"
format: html
editor: visual
---

## 

```{r}
library(readr)
library(ggplot2)
```

```{r}
SSMA <- read_csv("data/Students Social Media Addiction.csv")
```

```{r}
ggplot(SSMA, aes(x = Affects_Academic_Performance, fill = Academic_Level)) +
  geom_bar(position = "dodge") +
  labs(title = "Effect of Social Media on Academic Performance by Academic Level",
       x = "Affects Academic Performance",
       y = "Number of Students",
       fill = "Academic Level") +
  theme_minimal()


```

```{r}

SSMA$Academic_Level <- factor(
  SSMA$Academic_Level,
  levels = c("High School", "Undergraduate", "Graduate"),
  ordered = TRUE
)

ggplot(SSMA, aes(x = Academic_Level, y = Addicted_Score)) +
  geom_boxplot(fill = "skyblue") +
  labs(
    title = "Addicted Score by Academic Level",
    x = "Academic Level",
    y = "Addicted Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

```

## 



Methodology

This project investigates the impact of social media addiction on students using a publicly available dataset. Key variables include daily usage hours, addiction scores, academic performance impact, mental health ratings, interpersonal conflicts, and most used platforms. Data cleaning involved checking for missing values.

Descriptive and exploratory analyses were conducted to examine associations between addiction levels and outcomes. Boxplots, bar charts, and scatter plots were created to visualize variable relationships. A summary table was also generated to compare average usage hours and addiction scores by platform. 
