---
title: "The Impact of Social Media Addiction on Students"
format:
  revealjs:
    title-slide: false
    theme: simple
    transition: fade
---

# The Impact of Social Media Addiction on Students

:::{.columns}
::: {.column width="25%"}
*Nana Cheng*  
:::
::: {.column width="25%"}
*Kaiyi Ye*
:::
::: {.column width="25%"}
*Richa Anghan*
:::
::: {.column width="25%"}
*Yi Yang*
:::
::: 

2025-05-29

------------------------------------------------------------------------

## Introduction

-   Social media platforms are deeply embedded in students’ daily routines

-   Double-edged role:

    - <span style="font-size: 0.9em;">Support connection and self-expression</span> 
    - <span style="font-size: 0.9em;">Displace essential habits like sleep and study</span>

-   Authorities and researchers have raised concerns about social media in mental health issues

::: fragment
*How does social media addiction affect students’ academic performance, mental health, and relationships?*
:::

------------------------------------------------------------------------

## Dataset description

```{r}
#| echo: false
library(tidyverse)
library(dplyr)
library(readr)
```

```{r}
#| echo: false
addiction <- read.csv("data/Students Social Media Addiction.csv")
```

::::: columns
::: {.column width="45%"}
The dataset from Kaggle includes `r nrow(addiction)` student records and `r ncol(addiction)` variables on social media usage and its effects. Each row represents a student and covers variables like usage hours, addiction score, academic impact, sleep, and mental health.
:::

::: {.column width="55%"}
```{r}
#| label: tbl-variabletable
#| tbl-cap: Variable Names of the addiction Data

missing_data <- data.frame(
  Missing_Count = colSums(is.na(addiction))
)
missing_data
```
:::
:::::

------------------------------------------------------------------------

### Dataset description

```{r}
#| label: fig-Distribution
#| fig-cap: "Distribution of Addicted Scores"
#| eval: true

hist(addiction$Addicted_Score,
     breaks = 6,
     col = "skyblue",
     border = "black",
     main = "Distribution of Addicted Scores",
     xlab = "Addicted Score",
     ylab = "Frequency")
```

------------------------------------------------------------------------

## Method

Visual analyses were used to explore associations between social media addiction and student outcomes. These included:

-   A stacked bar chart showing the proportion of interpersonal conflicts by addiction score level.

-   A scatter plot with linear regression line to assess the relationship between daily usage hours and mental health scores.

-   A boxplot comparing addiction scores among students who reported academic impacts versus those who did not, stratified by academic level.

-   A summary table aggregating average usage hours and addiction scores by most used platform. ::::::
