---
title: "The Impact of Social Media Addiction on Students"
author:
  - Nana Cheng
  - Kaiyi Ye
  - Richa Anghan
  - Yi yang
date: 2025-05-29
format:
  revealjs:
    title-slide: false
    theme: simple
    transition: fade
---

## Introduction

-   Social media platforms are deeply embedded in students’ daily routines

-   Double-edged role:

    -   [Support connection and self-expression]{style="font-size: 0.8em;"}
    -   [Displace essential habits like sleep and study]{style="font-size: 0.8em;"}

- Authorities and researchers have raised concerns about social media in mental health issues  
  
::: fragment
*How does social media addiction affect students’ academic performance, mental health, and relationships?*
:::

<div style="font-size: 0.7em; color: gray; margin-top: 2em;">
Guardian, The. (2023). <i>Instagram lawsuit: Meta sued by dozens of US states over impact on youth mental health</i>. <i>The Guardian</i>.
</div>
------------------------------------------------------------------------

## Dataset description

```{r}
#| echo: false
library(tidyverse)
library(dplyr)
library(readr)
```

```{r}
#| echo: false
addiction <- read.csv("data/Students Social Media Addiction.csv")
```

The dataset from Kaggle includes `r nrow(addiction)` student records and `r ncol(addiction)` variables on social media usage and its effects. Each row represents a student and includes variables like:

```{r}
#| label: tbl-variabletable
#| tbl-cap: Variable Names of the addiction Data

missing_data <- data.frame(
  Missing_Count = colSums(is.na(addiction))
)
missing_data
```

------------------------------------------------------------------------

### Dataset description

```{r}
#| label: fig-Distribution
#| fig-cap: "Distribution of Addicted Scores"
#| eval: true

hist(addiction$Addicted_Score,
     breaks = 6,
     col = "skyblue",
     border = "black",
     main = "Distribution of Addicted Scores",
     xlab = "Addicted Score",
     ylab = "Frequency")
```

::: fragment
@fig-Distribution shows that most students report addiction scores between 6 and 8, with few below 4. This suggests a moderate to high level of perceived addiction and sufficient variation for further analysis.
:::

------------------------------------------------------------------------

## Method

-   A stacked bar chart showing the proportion of interpersonal conflicts by addiction score level.

-   A scatter plot with linear regression line to assess the relationship between daily usage hours and mental health scores.

-   A boxplot comparing addiction scores among students who reported academic impacts versus those who did not, stratified by academic level.

-   A summary table aggregating average usage hours and addiction scores by most used platform.

------------------------------------------------------------------------

## Analysis Part

### How does social media affect student

```{r}
library(ggplot2)
library(tidyverse)
library(kableExtra)
library(dplyr)
library(formattable)


```

```{r}
addiction <- read.csv("data/Students Social Media Addiction.csv")
```

:::::::: columns
:::: {.column width="33%"}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 10
ggplot(addiction, aes(x = factor(Addicted_Score), fill = factor(Conflicts_Over_Social_Media))) +
  geom_bar(position = "fill") +
  labs(
    title = "Conflicts vs Addiction",
    x = "Addiction Score",
    y = "Proportion",
    fill = "Conflicts Over Social Media"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",              
    legend.title = element_text(hjust = 0.5) 
  ) +
  guides(
    fill = guide_legend(title.position = "top", title.hjust = 0.5)  
  )

```

::: fragment
🧱 More addiction = more conflict High addiction scores strongly associate with interpersonal conflicts.
:::
::::

:::: {.column width="33%"}
```{r}

#| echo: false
#| fig-width: 10
#| fig-height: 10
ggplot(addiction, aes(x = Avg_Daily_Usage_Hours, y = Mental_Health_Score)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
```

::: fragment
📉 More screen time, worse mental health Daily use over 6 hours relates to low mental well-being.
:::
::::

:::
::::::::

:::

------------------------------------------------------------------------

## Platform Patterns & Academic Effects

:::::::: columns
:::: {.column width="40%"}
```{r}
platform_summary <- addiction %>%
  group_by(Most_Used_Platform) %>%
  summarise(
    Users = n(),
    Avg_Hours = round(mean(Avg_Daily_Usage_Hours, na.rm = TRUE), 1),
    Addiction_Score = round(mean(Addicted_Score, na.rm = TRUE), 1),
    .groups = 'drop'
  ) %>%
  arrange(desc(Avg_Hours)) %>%
  mutate(
    Rank = row_number(),
    Risk_Level = case_when(
      Avg_Hours >= 6 ~ "🔴 High",
      Avg_Hours >= 4 ~ "🟡 Medium", 
      TRUE ~ "🟢 Low"
    )
  ) %>%
  select(Rank, Most_Used_Platform, Users, Avg_Hours, Addiction_Score, Risk_Level) %>%
  slice_head(n = 5)
# Create improved table
platform_summary %>%
  kable(
    col.names = c("Rank", "Platform", "Users", "Daily Hours", "Addiction Score", "Risk Level"),
    align = c("c", "l", "c", "c", "c", "c"),
    caption = "📱 Top 5 Social Media Platforms by Daily Usage"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    font_size = 14,
    position = "center"
  ) %>%
  # Highlight top 3 platforms
  row_spec(1, bold = TRUE, color = "white", background = "#e74c3c") %>%
  row_spec(2, bold = TRUE, color = "white", background = "#f39c12") %>%
  row_spec(3, bold = TRUE, color = "white", background = "#f1c40f") %>%
  # Style header
  row_spec(0, bold = TRUE, color = "white", background = "#34495e", font_size = 16) %>%
  # Adjust column widths
  column_spec(1, bold = TRUE, width = "1.5cm") %>%
  column_spec(2, bold = TRUE, width = "4.5cm") %>%
  column_spec(3, width = "2.5cm") %>%
  column_spec(4, width = "2.5cm") %>%
  column_spec(5, width = "3cm") %>%
  column_spec(6, width = "3.5cm") %>% 
  scroll_box(width = "100%")


```

::: fragment
📱 WhatsApp & TikTok = high risk Top platforms show highest usage & addiction.
:::
::::

:::: {.column width="35%"}
```{r}
#| echo: false
#| fig-width: 4
#| fig-height: 4
ggplot(addiction, aes(x = Affects_Academic_Performance, y = Addicted_Score, fill = Affects_Academic_Performance)) +
  geom_boxplot() +
  facet_wrap(~ Academic_Level) +
  theme_minimal() +
  theme(
    legend.position = "bottom",        
    legend.direction = "horizontal",  
    legend.title = element_blank()    
  )

```

::: fragment
🎓 Addiction affects grades Across all academic levels, students reporting academic impact have higher addiction.
:::
::::


:::
::::::::

------------------------------------------------------------------------

## Conclusion

-   Social media significantly affects students' academic, mental, and social well-being.
-   Higher addiction scores correlate with poor academic performance.
-   Negative relationship found between daily usage hours and mental health.
-   WhatsApp, TikTok & Snapchat users show the highest usage and addiction scores.
-   Heavy engagement reduces focus, mental balance, and relationship quality.

------------------------------------------------------------------------

## Recommendations

-   **Set time limits** using digital well-being tools.
-   **Prioritize studies** before social media use.
-   Engage in purposeful, positive online interactions.
-   **Take breaks** from distressing platforms.
-   Invest in **offline social connections.**
-   **Raise awareness** on the effects of digital addiction.
-   Use social media for growth and learning.
-   Schools and universities should **offer digital wellness support.**
