---
title: "assignment 3"
format: html
editor: visual
---

## Methodology

This project investigates the impact of social media addiction on students using a publicly available dataset. Key variables include daily usage hours, addiction scores, academic performance impact, mental health ratings, interpersonal conflicts, and most used platforms. Data cleaning involved checking for missing values.

Descriptive and exploratory analyses were conducted to examine associations between addiction levels and outcomes. Boxplots, bar charts, and scatter plots were created to visualize variable relationships. A summary table was also generated to compare average usage hours and addiction scores by platform.

### Dataset Description

The student social media addiction data has `r nrow(addiction)` observations and `r ncol(addiction)` variables.

```{r}
library(tidyverse)
```

```{r}
addiction <- read.csv("data/Students Social Media Addiction.csv")
```

```{r}
head(addiction,2)
```

```{r}
#| label: tbl-variabletable
#| tbl-cap: Variable Names of the addiction Data

missing_data <- data.frame(
  Missing_Count = colSums(is.na(addiction))
)
missing_data
```

@tbl-variabletable

```{r}
#| label: fig-Distribution
#| fig-cap: "Distribution of Addicted Scores"
#| eval: true

hist(addiction$Addicted_Score,
     breaks = 6,
     col = "skyblue",
     border = "black",
     main = "Distribution of Addicted Scores",
     xlab = "Addicted Score",
     ylab = "Frequency")
```
